Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-02-16 paper/workshop/sparse_poly.tex
trying to make things more clear

\begin{abstract}
We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.

\end{abstract}

@@

Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
In this work we provide an algorithm for calculating polynomial features for a row vector $\vec{x}$ of a $N \times D$ matrix with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
The density of a vector is the percent of elements that are nonzero, so $0 \le d \le 1$.
The standard algorithm has time and space complexity $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.

\section{Algorithm}
The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination with repetition (without repetition for interaction features) of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, products resulting in zero will be common.
Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

The main idea behind our algorithm is to only compute products that do not involve zeros.
In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
We can therefore iterate over products of combinations with repetition of order $k$ of these columns for each row to calculate $k$-degree polynomial features.

While the idea is straightforward, there is yet one unaddressed caveat.
Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the following form:

\begin{equation}
(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
\forall (i_0, i_1, \dots, i_{k-1})
\end{equation}
such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$
and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
In general, and is equal to $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.

>>>>

\begin{abstract}
We provide an algorithm for polynomial feature expansion that can operate directly on sparse a matrix.
For a vector of dimensionality $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
\end{abstract}

@@

Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
In this work we provide an algorithm for calculating polynomial features for a vector of dimensionality $D$ and density $d$ with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
The density of a vector is the fraction of elements that are nonzero, so $0 \le d \le 1$.
The standard algorithm has time and space complexity $O(D^k)$, so the added factor of $d^k$ represents a significant reduction in complexity.
The algorithm does not require the densification of the vector, e.i. the vector remains in compressed sparse row form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.

\section{Algorithm}
The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of size $k$ with repetition (without repetition for interaction features) of features in $\vec{x}$.
This method does not exploit data sparsity and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, products resulting in zero will be common.
%If $\vec{x}$ has density $d$, the fraction of nonzero elements of the $k$-order expansion will be $\frac{\left(\binom{dD+k-1}{k}\right)}{\left(\binom{D+k-1}{k}\right)} = \frac{[(dD+k-1)(dD+k-2) \dots (dD+k)](D-1)!}{(dD-1)!} = $
If the vectors are stored in a sparse matrix format, the products resulting in zero are unnecessary to compute since zero is the default value in a sparse matrix format.

The main idea behind our algorithm is to leverage sparsity by only computing products that do not involve zeros.
In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
We can therefore iterate over products of combinations with repetition of order $k$ of these columns for each row to calculate $k$-degree polynomial features.

While the idea is straightforward, there is yet one unaddressed caveat;
Given a multiset of column indices whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the following form:

\begin{equation}
(i_0, i_1, \dots, i_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
\forall (i_0, i_1, \dots, i_{k-1})
\end{equation}
such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
where $(i_0, i_1, \dots, i_{k-1})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored,
and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
In general, $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
